The Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2013-11-12 papers/pair-correlation/paper.tex
a few edits to pair correlation paper Tested on collins. Took 1:10.41.

At its core, thermodynamic perturbation theory---sometimes referred to
as the cluster, or high-temperature expansion---is an expansion of the
free energy in powers of a small parameter, which typically is a
pairwise attractive interaction:
\begin{align}
  F &= F_0 + F_1 + \beta F_2 + \mathcal{O}(\beta^2)
\end{align}
where the terms $F_n$ are corrections to the free energy at order $n$
in the small interaction.  The first and largest term in this
expansion is
\begin{align}
  F_1 &= \tfrac12 \int g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|) d\rr_1d\rr_2
\end{align}
where $g^{(2)}_{HS}(\rr_1,\rr_2)$ is the pair distribution function of
the hard-sphere reference fluid, and $\Phi(r)$ is the pair potential.
In this paper, we will introduce an approximation for the hard-sphere
pair distribution function which is suitable for use in the creation
of classical density functionals based on a cluster expansion.

@@

Approaches that do not diverge use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel predicts the pair distribution function only at contact,
when the distance between $\rr_1$ and $\rr_2$ is a hard-sphere
diameter.

For these reasons Sokolowski and Fischer modified this approach to use
averages centered on the two points $\rr_1$ and $\rr_2$:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12},
  \tfrac12(\bar{n}(\rr_1)+\bar{n}(\rr_2))\right)
  \label{eq:sokolowski}
\end{align}
where
\begin{align}
  \bar{n}(\rr) = \frac{3}{4\pi (0.8\sigma)^3}\int n(\rr')\Theta(0.8\sigma - |\rr-\rr'|) d\rr'
\end{align}
is the density averaged over a sphere with diameter
$0.8\sigma$~\cite{sokolowski1992role}.  The value 0.8 in this formula
was arrived at by fitting to Monte Carlo simulation.  Although
Eq.~\ref{eq:sokolowski} has the advantages of only involving density
averages at the points at which the pair distribution function is
desired, cannot be written as a single-site convolution, since the
convolution kernel depends on both points.  This limits the accuracy.

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{\sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\subsection{Separable fit to the radial distribution function}
Figure~\ref{fig:radial-distribution} shows...

We here construct a separable fit to the radial distribution function
$g(r, g_\sigma)$, which is a function of the radius $r$ and its value
at contact $g_\sigma$.  We could convert this to a more conventional
function of the packing fraction $\eta$ by inverting the
Carnahan-Starling formula for $g_\sigma$
\begin{equation}
  g_\sigma = \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}
\end{equation}
but that would add an unnecessary step in the use of $g(r)$, since the
averaged value of $g_\sigma$ is what we can directly compute for the
inhomogeneous system.

@@

@@

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/triplet-correlation-pretty-contact-3.pdf}
  \caption{The triplet distribution function
    $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing fraction 0.3, plotted when
    $\rr_1$ and $\rr_2$ are in contact.  The color plot on the left
    shows the triplet distribution function, with white corresponding
    to the contact value of the \emph{pair} distribution function.
    The plot on the right shows the value of $g^{(3)}$ along the path
    shown as a dotted lin on the plot to the
    left.}\label{fig:triplet-contact-distribution-3}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/triplet-correlation-pretty-contact-4.pdf}
  \caption{The triplet distribution function
    $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing fraction 0.3, plotted when
    $\rr_1$ and $\rr_2$ are in contact.  The color plot on the left
    shows the triplet distribution function, with white corresponding
    to the contact value of the \emph{pair} distribution function.
    The plot on the right shows the value of $g^{(3)}$ along the path
    shown as a dotted lin on the plot to the
    left.}\label{fig:triplet-contact-distribution-4}
\end{figure}

@@

\section{How well does it work?}
Figure~\ref{fig:dadz} shows...

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/dadz-3-2.pdf}
  %\includegraphics[width=\columnwidth]{figs/dadz-3-3.pdf}
  \includegraphics[width=\columnwidth]{figs/dadz-2-2.pdf}
  \includegraphics[width=\columnwidth]{figs/dadz-square-well-3.pdf}
  %%\includegraphics[width=3cm]{figs/dadz-4-3}
  \caption{Plot of $\frac{da_1}{dz}$ near a hard
    wall. For what $\Phi$?}\label{fig:dadz}
\end{figure}

>>>>

At its core, thermodynamic perturbation theory---sometimes referred to
as the cluster, or high-temperature expansion---is an expansion of the
free energy in powers of a small parameter, which typically is a
pairwise attractive interaction:
\begin{align}
  F &= F_0 + F_1 + \beta F_2 + \mathcal{O}(\beta^2)
\end{align}
where the terms $F_n$ are corrections to the free energy at order $n$
in the small interaction.  The first and largest term in this
expansion is
\begin{align}
  F_1 &= \tfrac12 \iint g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|) d\rr_1d\rr_2
\end{align}
where $g^{(2)}_{HS}(\rr_1,\rr_2)$ is the pair distribution function of
the hard-sphere reference fluid, and $\Phi(r)$ is the pair potential.
In this paper, we will introduce an approximation for the hard-sphere
pair distribution function which is suitable for use in the creation
of classical density functionals based on a cluster expansion.

@@

Approaches that do not diverge use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  Moreover, the approach of Fischer and
Methfessel is intended to the pair distribution function only at contact,
when the distance between $\rr_1$ and $\rr_2$ is a hard-sphere
diameter.

For these reasons Sokolowski and Fischer modified this approach to use
averages centered on the two points $\rr_1$ and $\rr_2$:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12},
  \tfrac12(\bar{n}(\rr_1)+\bar{n}(\rr_2))\right)
  \label{eq:sokolowski}
\end{align}
where
\begin{align}
  \bar{n}(\rr) = \frac{3}{4\pi (0.8\sigma)^3}\int n(\rr')\Theta(0.8\sigma - |\rr-\rr'|) d\rr'
\end{align}
is the density averaged over a sphere with diameter
$0.8\sigma$~\cite{sokolowski1992role}.  The value 0.8 in this formula
was arrived at by fitting to Monte Carlo simulation.  Although
Eq.~\ref{eq:sokolowski} has the advantages of only involving density
averages at the points at which the pair distribution function is
desired, cannot be written as a single-site convolution, since the
convolution kernel depends on both points.

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{\sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\subsection{Separable fit to the radial distribution function}

We have constructed a separable fit to the radial distribution
function $g(r, g_\sigma)$, which is a function of the radius $r$ and
its value at contact $g_\sigma$.  We could convert this to a more
conventional function of the packing fraction $\eta$ by inverting the
Carnahan-Starling formula for $g_\sigma$
\begin{equation}
  g_\sigma = \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}
\end{equation}
but that would add an unnecessary step in the use of $g(r)$, since the
averaged value of $g_\sigma$ is what we can directly compute for the
inhomogeneous system~\cite{schulte2012using}.

@@

Figure~\ref{fig:radial-distribution} shows our function
$g(r,g_\sigma)$ plotted against the $g(r)$ found from simulation.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/triplet-correlation-pretty-contact-3.pdf}
  \caption{The triplet distribution function
    $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing fraction 0.3, plotted when
    $\rr_1$ and $\rr_2$ are in contact.  The color plot on the left
    shows the triplet distribution function, with white corresponding
    to the contact value of the \emph{pair} distribution function.
    The plot on the right shows the value of $g^{(3)}$ along the path
    shown as a dotted line on the plot to the
    left.}\label{fig:triplet-contact-distribution-3}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/triplet-correlation-pretty-contact-4.pdf}
  \caption{The triplet distribution function
    $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing fraction 0.3, plotted when
    $\rr_1$ and $\rr_2$ are in contact.  The color plot on the left
    shows the triplet distribution function, with white corresponding
    to the contact value of the \emph{pair} distribution function.
    The plot on the right shows the value of $g^{(3)}$ along the path
    shown as a dotted line on the plot to the
    left.}\label{fig:triplet-contact-distribution-4}
\end{figure}

@@

@@

\section{Performance in thermodynamic perturbation theory}

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/dadz-3-2.pdf}
  %\includegraphics[width=\columnwidth]{figs/dadz-3-3.pdf}
  %\includegraphics[width=\columnwidth]{figs/dadz-2-2.pdf}
  \includegraphics[width=\columnwidth]{figs/dadz-square-well-3.pdf}
  %%\includegraphics[width=3cm]{figs/dadz-4-3}
  \caption{Plot of $\frac{da_1}{dz}$ near a hard
    wall. For what $\Phi$?}\label{fig:dadz}
\end{figure}

The most relevant quantitative test of a pair distribution function is
how well it predicts the interaction energy due to a pair potential.
To this end, we have computed the error in the first term in a
high-temperature perturbation expansion $a_1$ for several pair
potentials.  In order to focus on effects at the interface, we have
defined a position-dependent pair interaction energy as
\begin{align}
  \frac{dF_1}{dz} &=
  \tfrac12 \int g^{(2)}_{HS}(\rr,\rr')n(\rr)n(\rr')\Phi(|\rr-\rr'|)
  d\rr'\, dxdy
\end{align}
which gives the contribution to the mean-field free energy due to
molecules located at position $z$.  We plot in Figure~\ref{fig:dadz}
this quantity for two pair potentials.

@@
